State of the Practice for Lattice Boltzmann Method Software

### 2022-02-09 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Updated the number of packages that use version control to be consistent throughout the doc

@@

Issue tracking is used in 23 software packages, 15 of which use GitHub or GitLab
to host their git repository, seven use email, and one (SunlightLB) uses
SourceForge. Most software packages that use GitHub/GitLab have most of their
issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50
percent of their issues closed. All of the top five overall ranked packages have
most of their issues closed. Alive packages (11 use issue tracking) have 64\% of
their issues closed, while dead packages (3 use issue tracking) have 71\% of
their issues closed. This information is presented in Table~\ref{gitrepodata}.
With respect to the host for the git repository, 13 packages use GitHub and two
(Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
uses CVS for issue tracking, and seven packages do not appear to use any issue
tracking system.

@@

\begin{table}
\begin{center}
\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
\toprule
\textbf{Common} & \textbf{Less Common} & \textbf{Rare} \\
\midrule
Authors / Developers List & Change Log / Release Notes & API Documentation\\
Issue Tracker & Design Documentation & Developer / Contributor Manual\\
Library Dependency List & Functional Spec. / Notes & FAQ / Forum\\
Installation Guide / Instructions & Performance Information & Verification and
Validation Plan\\
Theory Notes & Test Plan / Script / Cases & Video Guide
(including YouTube)\\
List of Related Publications & User Manual / Guide & Requirements Spec.\\
Makefile / Build File & Version Control & \\
README File & & \\
License & & \\
Tutorial & & \\
\bottomrule
\end{tabular}
\caption{Artifacts Present in LBM Packages, Classified by Frequency}
\label{artifactspresent}
\end{center}
\end{table}

The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
common artifacts, except Palabos does not have theory notes.  The top four
packages also have most of the less common artifacts. At the time of data
collection, only one of the four packages (Palabos) did not have a user manual
or guide, but there was a broken link on the package website indicating that
such an artifact might exist. This broken link was later fixed, but this is not
reflected in our data because it was not present at the time of data collection
(mid 2020). Despite the broken link, Palabos does have a detailed and
informative website. As far as we could observe, Ludwig does not appear to have
publicly visible design and OpenLB does not appear to use a version control
system. However, OpenLB's use of version numbers suggests that version control
may be used privately by the developers.

@@

\section{Comparison of Tool Usage to Research Software in General}
\label{Sec_CompareTools}

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts
\citep{ghezzi1991fundamentals}. Many tools are used by LBM software packages, as
summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
tools, dependencies, and project management tools.  In this section we answers
\rqref{RQ_CompareToolsProjMngmnt} by comparing the tool usage in LBM software to
the utilization in the research software community in general.

@@

Development tools support the development of end products, but do not become
part of them, unlike dependencies that remain in the application once it is
released \citep{ghezzi1991fundamentals}. Although not shown in
Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
Ludwig and Musubi) packages mentioned continuous integration tools, like Travis
CI. Code editors and compilers were explicitly noted to have been used by
several packages, and were likely used by all of them. One of the packages
(Ludwig) explicitly noted the use of proprietary unit testing code written in C.
Likewise, the use of proprietary code for verifying the correctness of output
was noted by one of the developers (pyLBM). Similar tools were likely used when
developing the other software packages.

For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
software packages use some sort of build automation tools, most commonly Make.
They all use various technical and domain specific libraries. Technical
libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
Open MPI, PyZMQ). Domain specific libraries include scientific computing
libraries (e.g.\ SciPy). Some used libraries will be missing, if they are not
explicitly stated in the artifacts we observed.

Many of the software packages that were assessed were developed by teams of two
or more people. Their work needed to be coordinated and managed.
Table~\ref{tbl_tools} shows the types of project management tools that were
explicitly noted in the artifacts, web-pages, or interviews with the developers.
As with development tools and dependencies, it is possible that other types of
project management tools are used, but they were not visible in the artifacts to
which we had access.  Collaboration tools are noted as being used when
developing the software projects. Most often email and video conferencing is
used. Project management software was not explicitly mentioned, but it is
possible that some of the projects use such software. Many of the projects are
located on GitHub, and its developers use the platform to help manage their
projects, especially bug related issues. Most of the projects appear to use
change tracking and version control tools. \wss{**Version Control**} \wss{This
contrasts with work from a little over 10 years ago that shows version control
was only used in 50\% of research software projects \citep{Nguyen-HoanEtAl2010}.
At that time \citet{Nguyen-HoanEtAl2010} noted an increase from previous levels
of version control software use.  Anecdotally this confirms our observations.}
They often use GitHub or GitLab for this. One package (SunlightLB) uses CVS.
Document generation tools are mentioned in the artifacts of 12 of the projects.
The tools Sphinx and Doxygen are explicitly used in this capacity. \wss{look at
what the other papers say are commonly used tools and bring that information
into this section.}

\wss{Beyond software carpenty has data on the increasing use of version control.}

>>>>

\subsection{Maintainability} \label{Sec_Maintainability}

@@

Issue tracking is used in 23 software packages, 15 of which use GitHub or GitLab
to host their git repository, seven use email, and one (SunlightLB) uses
SourceForge. Most software packages that use GitHub/GitLab have most of their
issues closed, and only three (laboetie, lettuce, Sailfish) have less than 50
percent of their issues closed. All of the top five overall ranked packages have
most of their issues closed. Alive packages (11 use issue tracking) have 64\% of
their issues closed, while dead packages (3 use issue tracking) have 71\% of
their issues closed. This information is presented in Table~\ref{gitrepodata}.
With respect to the host for the git repository, 13 packages use GitHub and two
(Palabos, waLBerla) use GitLab. Of the other packages, one package (SunlightLB)
uses CVS for issue tracking and version control, and seven packages do not
appear to use any issue tracking system.

@@

\begin{table}
\begin{center}
\begin{tabular}{ p{5.3 cm} p{4.9 cm} p{5 cm}}
\toprule
\textbf{Common} & \textbf{Less Common} & \textbf{Rare} \\
\midrule
Authors / Developers List & Change Log / Release Notes & API Documentation\\
Issue Tracker & Design Documentation & Developer / Contributor Manual\\
Library Dependency List & Functional Spec. / Notes & FAQ / Forum\\
Installation Guide / Instructions & Performance Information & Verification and
Validation Plan\\
Theory Notes & Test Plan / Script / Cases & Video Guide
(including YouTube)\\
List of Related Publications & User Manual / Guide & Requirements Spec.\\
Makefile / Build File &  & \\
README File & & \\
License & & \\
Tutorial & & \\
Version Control & & \\
\bottomrule
\end{tabular}
\caption{Artifacts Present in LBM Packages, Classified by Frequency}
\label{artifactspresent}
\end{center}
\end{table}

The top four packages, ESPREsSo, Ludwig, Palabos, and OpenLB, have all of the
common artifacts, except Palabos does not have theory notes and only three of
them (ESPREsSo, Ludwig, Palabos) appear to use a version control system.  The
top four packages also have most of the less common artifacts. At the time of
data collection, only one of the four packages (Palabos) did not have a user
manual or guide, but there was a broken link on the package website indicating
that such an artifact might exist. This broken link was later fixed, but this is
not reflected in our data because it was not present at the time of data
collection (mid 2020). Despite the broken link, Palabos does have a detailed and
informative website. As far as we could observe, Ludwig does not appear to have
publicly visible design and OpenLB does not appear to use a version control
system. However, OpenLB's use of version numbers suggests that version control
may be used privately by the developers.

@@

\section{Comparison of Tool Usage Between LBM and Other Research Software}
\label{Sec_CompareTools}

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts \citep[p.\
501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
tools, dependencies, and project management tools.  In this section we answers
\rqref{RQ_CompareToolsProjMngmnt} by comparing the tool usage in LBM to
their utilization in the research software community in general.

@@

Development tools support the development of end products, but do not become
part of them, unlike dependencies that remain in the application once it is
released \citep[p.\ 506]{GhezziEtAl2003}. Although not shown in
Table~\ref{tbl_tools}, debuggers were also likely used.  Only three (ESPResSo,
Ludwig and Musubi) packages mentioned continuous integration tools, like Travis
CI. Code editors and compilers were explicitly noted to have been used by
several packages, and were likely used by all of them. One of the packages
(Ludwig) explicitly noted the use of proprietary unit testing code written in C.
Likewise, the use of proprietary code for verifying the correctness of output
was noted by one of the developers (pyLBM). Similar tools were likely used when
developing the other software packages.

For the dependency tools (Table~\ref{tbl_tools}), we observed that most of the
software packages use some sort of build automation tools, most commonly Make.
They all use various technical and domain specific libraries. Technical
libraries include visualization (e.g.\ Matplotlib, ParaView, Pygame, VTK), data
analysis (e.g.\ Anaconda, Torch), and message passing libraries (e.g.\ MPICH,
Open MPI, PyZMQ). Domain specific libraries include scientific computing
libraries (e.g.\ SciPy).

Many of the software packages that were assessed were developed by teams of two
or more people. Their work needed to be coordinated and managed.
Table~\ref{tbl_tools} shows the types of project management tools that were
explicitly noted in the artifacts, web-pages, or interviews with the developers.
As with development tools and dependencies, it is possible that other types of
project management tools are used, but they were not visible in the artifacts to
which we had access.  Collaboration tools, most often email and video, are noted
as being used when developing the software projects. Project management software
was not explicitly mentioned, but it is possible that some of the projects use
such software. Many of the projects are located on GitHub, where the developers
use the platform to help manage their projects, especially bug related issues.
Most of the projects appear to use change tracking and version control tools.
They often use GitHub or GitLab for this. One package (SunlightLB) uses CVS.
Document generation tools are mentioned in 12 of the 24 projects. The tools
Sphinx and Doxygen are explicitly used in this capacity.

The poor adoption of version control tools that Wilson lamented in 2006
\citep{Wilson2006} has greatly improved in the intervening years.  From
Section~\ref{Sec_Maintainability} 67\% of LBM packages use version control
(GitHub, GitLab and CVS).  The proliferation of version control tools for LBM
matches the increase in the broader research software community.  A little over
10 years ago (in 2010) version control was estimated to be used in only 50\% of
research software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels.  A
survey in 2018 shows 81\% of developers use a version control system
\citep{AlNoamanyAndBorghi2018}.  This matches with \citet{Smith2018} shows
active projects in mesh generation, geographic information systems and
statistical software for psychiatry increasing from 75\%, 89\% and 17\%
(respectively) to 100\%, 95\% and 100\% (respectively) over a four year period
ending in 2018.  (For completeness the same study showed a decrease in version
control tool usage for seismology software over the same time period (from 41\%
down to 36\%).  Almost every software guide cited in
Section~\ref{Sec_CompareArtifacts} mentions the advice to use version control.
